The role of lithic bipolar technology in Western Iberia’s Upper Paleolithic: the case of Vale Boi (southern Portugal)

### 2018-12-09 analysis/paper/paper.Rmd
data dictionaries altered

output: 
    bookdown::word_document2: #substitute by pdf_document2 or hmtl_document2 as necessary
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-socpsych-author-date.csl"
---

@@

Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*). Their distinction, however, has not always been consensual [@hayden_confusion_1980]. Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact. The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool. Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g. @hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].

Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been largely debated [e.g. @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963]. The issue surrounding this problem lies within the functional equifinality of these artifacts. Contrary to bipolar cores that are most of the times easily identified as cores, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard raw materials (e.g. bone, ivory, antler); and (2) as cores for the extraction of chips and small bladelets [@brantingham_early_2004]. Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g. @hiscock_dynamics_2015].

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of portuguese paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces. Most references broadly interpret these as bipolar cores for the extraction of small bladelets and chips to be  used in composite tools [e.g. @zilhao_o_1997], or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997], for example, argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the portuguese early Neolithic.

While we agree that some of the elements might have been used as cores, these interpretations seem overly simplistic, and the authors seem to have little to no evidence to back their claims other than the inverse relationship mentioned above. This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale. Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation and, thus, no direct association between these tools and the exploitation of hard-organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.

@@

The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)). The site can be found in a small valley following a river north-south for 2 km until it reaches the Atlantic Ocean. Archaeological deposits occupy an estimated area of over 10 000 sq. meters across a stepped slope marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge:_2012; @bicho_paleolithique_2010].

@@

The Terrace area is located in the lower part of the hill. In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations [@bicho_ledge:_2012; @cascalheira_cross-scale_2017]. From within the lower levels of this area, an Early Gravettian occupation was discovered, dated to c. 32 ka cal BP, being one of the earliest radiocarbon dates for anatomically modern humans in Southern Iberia [@bicho_ecodynamics_2013; @marreiros_lithic_2015].

@@

Finally, the Slope section, from where the assemblage here presented is coming from, is composed of a series of excavation areas opened across the mid-hill sector of the site. These areas exhibit heterogeneous conditions in terms of site formation processes and archaeological preservation [@manne_intensive_2012], but all revealed the presence of occupations attributed to the Gravettian, Proto-Solutrean, Solutrean and Magdalenian. Like in the previous areas, remains are well preserved, and high frequencies of lithic artifacts, malacological and mammalogical fauna and bone tools were recovered. No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have mostly functioned as a midden deposit [@bicho_ledge:_2012; @bicho_paleolithique_2010].

@@

Chert is the most abundant knapped rock type in Vale Boi. Throughout the several occupations, chert was considered the best rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz. While it can be said that chert exploration strategies were more elaborate, the most common strategies were still simple unidirectional or bidirectional reduction sequences, producing mostly flakes [@marreiros_flake_2012]. Elongated products are found at low frequencies across all chronologies. When present, retouched tools are mostly notches, denticulates, end-scrapers, and scaled pieces. One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures [@pereira_territory_2016].

Two distinct types of quartz were identified at the site. The first is a thick grain, low-quality type, mostly inadequate for knapping. Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities [@bicho_ledge:_2012; @manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. The second type of quartz is characterized by finer-grain small pebbles with yellowish cortex. This type of quartz was knapped using rather simple strategies, mostly for small flake production, which in turn were used to produce simple and versatile tools.

Greywacke is the third most frequent raw material at Vale Boi, showing up in the site mostly in the form of large slabs, in which the identification of concavely shaped impact marks is thought to result from their use as anvils. The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations [@bicho_ledge:_2012; @manne_intensive_2012]. Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies [@cascalheira_influencia_2013; @marreiros_as_2009].

@@

We define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends. These traces are caused by direct percussion at one end and subsequent crushing of the other because it is rested on a hard surface, and can be bifacially or unifacially distributed. As will become clear from our results, some of the analyzed artifacts did not present splitering/crushing traces in opposed platforms, but we still classified them as scaled pieces. This detail has been previously referred by Villa et al. [-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platforms.

Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes and (2) morpho-functional attributes. In the first group, a series of variables traditionally used in lithic studies [e.g. @andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded (see Supplementary Materials for full list of variables), aiming to characterize patterns of blank choice for the application of bipolar technology. For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et. al [-@fischer_macro_1984] was adopted (see Supplementary Materials for full list of variables). Studies by Sano [-@sano_functional_2012], and Gibaja et al. [-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials). This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013]. Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our supplemental online material (SOM) at https://dx.doi.org/10.17605/OSF.IO/WPXGH To produce those files we followed the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].

@@

As previously mentioned, concerning technological data, one of our main objectives was to characterize the choice of blanks involved in bipolar technology. Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site. Across all techno-complexes, blanks used were either flakes or unclassifiable fragments, but the choice seems to be different for each raw material. For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)). In other periods, flakes were the preferred type of blank (Tables \@ref(tab:technosolutrean), \@ref(tab:technoproto) and \@ref(tab:technomagdalenian)). Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks. The small sample of chalcedony blanks is exclusively represented by flakes.

@@

Other characteristics of the assemblage are the low frequency of retouch (n = `r techno_data %>% filter(RetouchedBlank == "Yes") %>% summarise(n())`), of fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and of the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`). Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper. In a very small number of cases, when striking platforms are present, these are mostly flat. The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice. The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis. In fact, when identifiable, technological and typological axes coincide in `r inline.perc(techno_data, AxisCoincident, "Yes")`% of the cases. The longitudinal sections show a large variability of shapes, independent of raw materials or chronologies. Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.

@@

# Set variables
var_list <- c("Blank","CrossSection","Profile","Blankshape","Cortex","ButtType","Profile")

@@

techno_data <- techno_data %>% 
  mutate(Area = AxisWidth*AxisLength)

@@

aov_thickness$p.value

@@

The patterns of use and rotation of damaged platforms seem to be the same across all chronologies. In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms. On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)). This seems to indicate different strategies of curation for coarse and fine grain raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity. However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thickness of the pieces comprising each group. This seems to attest that the use of several axis in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.

@@

@@

In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable. A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = `r round(chi_arrangement$chi_test$p.value, 3)`, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`). The effect size statistic (Phi) suggests, however, a medium practical significance for both variables.

@@

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g. @villa_neandertals_2018]. A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive. Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits. It is in the context of this latter point that bipolar technology may have played a major role. In Vale Boi, since the earliest occupations at c. 32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the aforementioned grease/marrow obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017]. For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture. In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision. The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.

@@

Bipolar technology is quite common in Upper Paleolithic contexts of Western Iberia, but current interpretations are based on empirical observations, rather than on more solid analytical evidence. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be for bipolar knapping, in other sites the picture may be different.

@@

# Acknowledgements
We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade. These are Archaeological Association of Algarve, Fundaç~ao para a Ciência e Tecnologia, Instituto Português de Arqueologia, National Geographic Society, and Wenner-Gren Foundation for Anthropological Research.

>>>>

output: 
    bookdown::word_document2:
      fig_caption: yes
      reference_docx: "../templates/template2.docx" 
bibliography: ScaledPiecesVB.bib
csl: "../templates/springer-socpsych-author-date.csl"
---

@@

Bipolar technology is generically classified into two types of lithic artifacts - bipolar cores and scaled pieces (also known as splintered pieces or *pièce esquilée*), this latter being the main focus of this paper. Their distinction, however, has not always been consensual [@hayden_confusion_1980]. Besides the fact that bipolar cores are often confused or lumped with scaled pieces [@villa_neandertals_2018;@de_la_pena_sobre_2011], the definition of scaled piece has suffered significant changes ever since its initial identification in the early 20th century and, to this day, there still seems to be no clear definition for this type of artifact. The first definition of scaled pieces was proposed by Bardon and Bouyssonie [-@bardon_outils_1906], describing them as a result of bipolar knapping through direct percussion, with the “core” rested in a hard surface, originating splintering in both ends of the tool. Since then, several other definitions were adopted and adapted by researchers for a very diverse set of contexts across the world [e.g. @hayden_confusion_1980; @macdonald_debert:_1985; @octobon_contribution_1938; @shott_bipolar_1989; @sonneville-bordes_lexique_1956].

Beyond the classification debate, the biggest issue with this type of tools comes from a functional standpoint, a problem that has also been largely debated [e.g. @binford_indian_1963; @de_la_pena_sobre_2011; @flood_moth_1980; @igreja_functional_2013; @leblanc_wedges_1992; @lucas_les_2004; @shott_bipolar_1999; @tixier_typologie_1963]. The issue surrounding this problem lies within the functional equifinality of these artifacts. Contrary to bipolar cores that are exclusively part of a technological reduction sequence with the goal of blank extraction, scaled pieces have been associated with two distinct types of activities: (1) as intermediate pieces or wedges for working hard organic raw materials (e.g. bone, ivory, antler); and (2) as cores for the extraction of chips, small flakes and bladelets [@brantingham_early_2004]. Although largely debated, the ambiguity of this classification has, however, been ignored in some of the most recent literature, with some authors not acknowledging that bipolar evidence in stone tools might also result from other activities other than only reduction strategies [see e.g. @hiscock_dynamics_2015].

@@

In the case of the westernmost regions of Iberia, scaled pieces are commonly found in archaeological contexts ranging from the Upper Paleolithic to the Neolithic [e.g. @bicho_technological_2000; @carvalho_talhe_1998; @zilhao_o_1997]. While scarce, the majority of Portuguese Paleolithic studies [e.g. @almeida_terminal_2000; @goncalves_variabilite_2012] have not presented, so far, any context-specific interpretations for the presence of scaled pieces. Most references broadly interpret these as bipolar cores for the extraction of small flakes, bladelets and chips to be used in composite tools [e.g. @zilhao_o_1997], or as intermediate pieces for working hard materials [@gibaja_tecnologiy_2007; @marreiros_as_2009]. Zilhão [-@zilhao_o_1997], for example, argues that the presence of scaled pieces throughout the Upper Paleolithic sequence of central Portugal is inversely proportional to the presence of “carinated cores”, and thus it is likely that the former should have worked as a flexible substitute for the latter. Carvalho [-@carvalho_talhe_1998] agrees with this interpretation and considers it also valid for the Portuguese early Neolithic.

While we agree that some of the elements might have been used as cores, these interpretations seem simplistic, and the authors seem to have little to no analytical evidence to back their hypothesis, other than the inverse relationship mentioned above. This results, in part, from the lack of comprehensive studies regarding this type of artifacts at a regional scale. Additionally, the contexts from where most of these artifacts were recovered did not have good organic preservation or dedicated use-wear studies and, thus, no direct association between these tools and the exploitation of hard organic raw materials is possible, and, in face of the nonexistence of any absolute dates, their precise chronological attribution is also, most of the times, unreliable.

@@

The archaeological site of Vale Boi is located in the western edge of the Algarve region (southern Portugal) (Figure \@ref(fig:VBLocation)). The site can be found in a small valley following a river north-south for 2 km, until it reaches the Atlantic Ocean. Archaeological deposits occupy an estimated area of over 10 000 sq. meters across a stepped slope, marked at the top by a 10 meter-high limestone cliff face [@bicho_ledge_2012; @bicho_paleolithique_2010].

@@

The Terrace area is located in the lower part of the hill. In this area, the longest archaeological sequence of the site can be found, including the complete Upper Paleolithic sequence but also three Holocene horizons, corresponding to Neolithic, Mesolithic and Epipaleolithic occupations [@bicho_ledge_2012; @cascalheira_cross-scale_2017]. From within the lower levels of this area, an Early Gravettian occupation was discovered, dated to c. 32 ka cal BP, being one of the earliest radiocarbon dates for anatomically modern humans in Southern Iberia [@bicho_ecodynamics_2013; @marreiros_lithic_2015].

@@

Finally, the Slope section, from where the assemblage here presented is coming from, is composed of a series of excavation areas opened across the mid-hill sector of the site. These areas exhibit heterogeneous conditions in terms of site formation processes and archaeological preservation [@manne_intensive_2012], but all revealed the presence of occupations attributed to the Gravettian, Proto-Solutrean, Solutrean and Magdalenian. Like in the previous areas, remains are well preserved, and high frequencies of lithic artifacts, malacological and mammalogical fauna and bone tools were recovered. No habitation features were identified in this area and based on the conditions and type of artifacts found, it has been suggested that this area would have mostly functioned as a midden deposit [@bicho_ledge_2012; @bicho_paleolithique_2010].

@@

Chert is the most abundant knapped rock type in Vale Boi. Throughout the several occupations, chert was considered the preferred rock type for knapping, which is evident by the way it was explored with much more complex strategies than either greywacke or quartz. While it can be said that chert exploration strategies were more elaborate, the most common strategies were still simple unidirectional or bidirectional reduction sequences, producing mostly flakes [@marreiros_flake_2012]. Elongated products are found at low frequencies across all chronologies. When present, retouched tools are mostly notches, denticulates, end-scrapers, and, although not retouched artifacts *per* *se* scaled pieces. One possible reason for these simplistic strategies seems to be an overall low knapping quality, since nodules are quite small, and frequently show a high degree of tectonically derived fractures [@pereira_territory_2016].

Two distinct types of quartz were identified at the site. The first is a thick grain, low-quality type, mostly inadequate for knapping. Still, this kind of quartz is present in large quantities and is most likely associated with stone boiling and grease rendering activities [@bicho_ledge_2012; @manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. The second type of quartz is characterized by finer-grain small pebbles with yellowish cortex. This type of quartz was knapped using rather simple strategies, mostly for small flake production, which in turn were used to produce simple and versatile tools.

Greywacke is the third most frequent raw material at Vale Boi, showing up in the site mostly in the form of large slabs, in which the identification of concavely shaped impact marks is thought to result from their use as anvils. The importance of greywacke anvils is indicated by its high frequency, with hundreds of slabs found in many levels throughout all occupations [@bicho_ledge_2012; @manne_intensive_2012]. Greywacke was also knapped for flake extraction, using expedient unidirectional reduction strategies [@cascalheira_influencia_2013; @marreiros_as_2009].

@@

Based on previous works [e.g. @hayden_confusion_1980] here we define scaled pieces as artifacts of variable size and morphology, showing traces of crushing and splitering of edges at opposite ends, caused by direct percussion at one end, and subsequent crushing of the other for being rested on a hard surface. Scaled pieces present always two opposite surfaces, just like a flake or blade would. Crushing traces can be bifacially or unifacially distributed. Some scaled pieces do not present crushing in opposed platforms, but are still classified as scaled pieces here. This detail has been previously referred by Villa et al. [-@villa_neandertals_2018] who noted that some edges may present a flat (instead of intensively shattered) platform. Our criteria does not include a particular type of blank because, as we will show in later sections of the paper, these artifacts can originate from flakes, blades or debris.

On the contrary, following mostly Hayden [-@hayden_confusion_1980] and Leaf [-@leaf_variation_1979], bipolar cores are not marked by bifacially opposed surfaces. Their shape is more blocky and angular, showing evidence of, at least, one flaking surface with two opposed platforms (the striking platform and the base) with typical crushing and flake removal on one or both ends. The striking platform is the surface that is struck with a hammerstone to produce blanks. Tipically, it exhibits little crushing except near the point of impact. Large flake scars tend to originate at the striking platform. The base is the surface that rests on the anvil, from where small flake scars can also originate.

All lithic artifacts recovered from the Slope area of Vale Boi that fit the scaled piece definition presented below were considered for this study, independent of raw material or technological class.

Attribute analysis was split into two main groups, each corresponding to two distinct types of features: (1) technological attributes (Table \@ref(tab:techno_attributes)) and (2) morpho-functional attributes (Table \@ref(tab:morpho_attributes)). In the first group, a series of variables traditionally used in lithic studies [e.g. @andrefsky_jr_lithics:_2005; @inizan_technology_1999] were recorded, aiming to characterize patterns of blank choice for the application of bipolar technology. For the second group of variables, a macroscopic approach building upon the work of de la Peña [-@de_la_pena_sobre_2011] and Fischer et. al [-@fischer_macro_1984] was adopted. Studies by Sano [-@sano_functional_2012], and Gibaja et al. [-@gibaja_tecnologiy_2007] indicate that, since use-wear traces are formed before the splintering, the removal of small chips removes most of the polishes and use-wear traces left by the contact with the static element (either a stone anvil or a hard-organic materials). This, together with the large presence of quartz artifacts in our sample, prevented us to pursue a microscopic use-wear approach for this study in particular.

```{r techno_attributes, echo=FALSE}

techno_attributes <- read_csv("../data/raw_data/technological_attributes.csv")

morpho_attributes[is.na(techno_attributes)] <- " "

knitr::kable(techno_attributes, caption = "Technological attributes used for the analysis of scaled pieces included in this study.")

```{r morpho_attributes, echo=FALSE}

morpho_attributes <- read_csv("../data/raw_data/functional_attributes.csv")

morpho_attributes[is.na(morpho_attributes)] <- " "

knitr::kable(morpho_attributes, caption = "Morpho-functional attributes used for the analysis of each damaged platform of scaled pieces included in this study. *Adapted from Gonzalez-Urquijo and Ibanez-Estévez [-@gonzalez-urquijo_metodologialisis_1994]. **Adapted from de la Peña [-@de_la_pena_sobre_2011].")

@@

All analyses and data processing were accomplished in R (version `r getRversion()`) [@r_core_team_r:_2013]. Following recent concerns on the reproducibility of archaeological analysis we include the entire R code used for all the analysis and visualizations contained in this paper in our online research compendiumm at https://dx.doi.org/10.17605/OSF.IO/WPXGH. To produce those files we followed the procedures described by Marwick et al. [-@marwick_packaging_2017] for the creation of research compendiums to enhance the reproducibility of research. The files provided contain all the raw data used in our analysis as well as a custom R package [@wickham_r_2015] holding the code use for all analysis and to produce all tables and figures. To enable maximum re-use, our code is released under the MIT license, our data as CC-0, and our figures as CC-BY, [for more information see @marwick_computational_2016].

@@

As previously mentioned, concerning technological data, one of our main objectives was to characterize the choice of blanks involved in bipolar technology. Overall, the technological analysis revealed some trends that lasted throughout the Upper Paleolithic, in agreement with the general patterns of lithic technology recorded for the site. Across all techno-complexes, blanks used were either flakes (based on the recognition of dorsal and ventral surfaces) or unclassifiable fragments, but the choice seems to be different for each raw material. For quartz, in the Gravettian assemblage, the blank types are almost equally split between flakes and unclassifiable pieces (Table \@ref(tab:technogravettian)). In other periods, flakes were the preferred type of blank (Tables \@ref(tab:technosolutrean), \@ref(tab:technoproto) and \@ref(tab:technomagdalenian)). Regarding chert, in every occupation flakes dominate the assemblages, followed by a reduced number (n = `r techno_data %>% filter(RawMaterial == "Chert" & Blank == "Non_identifiable") %>% summarise(n())`) of unclassifiable blanks. The small sample of chalcedony blanks is exclusively represented by flakes.

@@

Other characteristics of the assemblage are the low frequency of retouch found in the non-damaged edges (n = `r techno_data %>% filter(Retouch == "Yes") %>% summarise(n())`), fire alterations (n = `r techno_data %>% filter(Fire != "No_Traces") %>% summarise(n())`), and the presence of original striking platforms (n = `r techno_data %>% filter(ButtType != "Absent") %>% summarise(n())`). Still, when present, retouch is located in the lateral part of the artifacts, similar to what would define a side-scraper. In a very small number of cases, when striking platforms are present, these are mostly flat. The absence of the original blank platforms is to be expected in this type of artifact, mostly due to the functional use of the pieces, and consequent removal of the platform, rather than an actual choice. The large absence of platforms may, also, be the result of the use of the technological axis as main functional axis. In fact, when identifiable, technological and typological axes coincide in `r inline.perc(techno_data, AxisCoincident, "Yes")`% of the cases. The longitudinal sections show a large variability of shapes, independent of raw materials or chronologies. Similarly, the dorsal pattern of previous removals was rarely identified, although this, like with the case of platform absence, might occur due to the functional stigmas and be dependent on the intensity of use for each artifact.

@@

# Set variables
var_list <- c("Blank","AxisCoincident","CrossSection","SideSection","Profile","BlankShape","Cortex","ButtType","Retouch","Fire")

@@

techno_data <- techno_data %>% 
  mutate(Area = TypologicalAxisWidth*TypologicalAxisLength)

@@

@@

The patterns of use and rotation of damaged platforms seem to be the same across all chronologies. In every assemblage quartz pieces were exclusively used in one single axis, exhibiting only two damaged platforms. On the other end, a small number of chert and chalcedony artifacts (n = `r techno_data %>% filter(RawMaterial != "Quartz" & DamagedPlatforms == "3" | DamagedPlatforms == "4_or_more") %>% summarize(n())`) exhibit multiple functional axis, with three or four damaged platforms (Figure \@ref(fig:nplatforms)). This seems to indicate different strategies of curation for coarse (quartz) and fine grain (chert and chalcedony) raw materials, with fine grain materials being rotated when the first used axis becomes too small and/or the edges get useless for that specific activity. However, when plotted against metric data (Figure \@ref(fig:plotbyplatforms)), results for the Area variable reveal that pieces with four damaged platforms are among the largest in the whole assemblages, and that there is not a visible difference between the thicknesses of the pieces comprising each group. This seems to attest that the use of several axes in the same piece is not correlated with curation occuring in later phases of artifact use but instead to a probable difference in raw material performance.

@@

chi_arrangement$chi_test

@@

In three (Scar Arrangement, Scar Extension and Edge Delineation) of the six represented variables, the category 'Other' is one of the most frequent (above 35%) for both quartz and chert, revealing a very high variability for the combination of attributes within each variable. A chi-square test with modified alpha level (Bonferroni correction) to adjust for multiple testing and reduce type-I error, shows significant differences in the quartz and chert Scar Arrangement categories (*X^2^* [`r chi_arrangement$chi_test$parameter`, *N* = `r sum(chi_arrangement$chi_test$observed)`] = `r  round(chi_arrangement$chi_test$statistic, 2)`, *p* = < 0.001, *phi* = `r round(chi_arrangement$chi_size_effect$phi, 2)`), and in Scar Faciality categories (*X^2^* [`r chi_faciality$chi_test$parameter`, *N* = `r sum(chi_faciality$chi_test$observed)`] = `r  round(chi_faciality$chi_test$statistic,2)`, *p* = `r round(chi_faciality$chi_test$p.value, 3)`, *phi* = `r round(chi_faciality$chi_size_effect$phi, 2)`). The effect size statistic (Phi) suggests, however, a medium practical significance for both variables.

@@

Bipolar technology clearly had an important role on the adaptive systems of the first modern humans in Western Iberia, as well as in other european regions [see e.g. @villa_neandertals_2018; @de_la_pena_sobre_2011; @sano_functional_2012; @zilhao_o_1997]. In Europe there is a considerable rise in the use of bipolar technologies after the Middle to Upper Paleolithic transition, and some authors consider scaled pieces as one of the most common lithic morphotypes in european Upper Paleolithic assemblages [@de_la_pena_sobre_2011]. A rise in bipolar technologies cannot be dissociated from the diverse set of factors that made Anatomically Modern Humans thrive. Changes in mobility patterns [@shott_diversity_2015], or the development of a "generalist specialist" ecological approach [@roberts_defining_2018], with particular emphasis on the diversification and intensification of resources use, are among some of those traits. It is in the context of this latter point that bipolar technology may have played a major role. In Vale Boi, since the earliest occupations at c. 32 ka cal BP, scaled pieces show up in the archaeological record connected to evidence related to an intensification and diversification of resource exploration, particularly the aforementioned grease/marrow obtention techniques. These have been shown by the constant presence of impact fractures in ungulate bones, low percentage of bone areas related to higher amounts of grease, and the presence of red deer bone fragments from bones with considerably higher marrow and grease contents [@manne_early_2014; @manne_intensive_2012; @manne_upper_2010; @manne_vale_2009]. Other indicators are both the constant presence of thermally altered quartz fragments that might be linked to stone boiling activities, and a large number of greywacke slabs with impact marks revealing their use as anvils [@cascalheira_cross-scale_2017]. For carcass processing and bone marrow extraction the use of scaled pieces would allow for a better control of bone fracture. In fact, this should be significantly better than using a hard hammer directly on the bone, since the latter technique may either over fracture the bone or even crush it due to the lack of precision. The use of a wedge for these activities provides more control, avoiding complete crushing of the bone, making it much easier to cleanly extract bone marrow.

@@

Bipolar technology is quite common in Upper Paleolithic contexts of Western Iberia, but current interpretations are based on empirical observations, rather than on more solid analytical evidence. This, of course, stems in part from the lack of dedicated studies on bipolar technology in the region, an issue that can be also argued to be true for all of the European Upper Paleolithic. We argue that all factors mentioned in this study must be considered while interpreting each site in this region. While in Vale Boi there is clear evidence that the use of these artifacts would not be primarly for bipolar knapping, in other sites the picture may be different.

@@

# Acknowledgements
We would like to thank a diversity of institutions that provided funding to the work in Vale Boi in the last decade. These are Fundação para a Ciência e Tecnologia (Grants PTDC/HAH/64184/2006, SFRH/BD/65527/2009 and PTDC/HIS-ARQ/117540/2010), National Geographic Society (Grant #8045-06), and Wenner-Gren Foundation for Anthropological Research (Grant #8290). Thanks also to the three anonymous reviewers for their valuable comments that helped to improve the content of the paper.
